EP1162 Preliminary Analysis

### 2020-03-21 analysis/heightModel.Rmd
more updates

****
### Can we pool species and measurement years?

@@

```{r test for species and year effect}

# Assign data
x<-
   dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
    drop_na(Height) # remove trees where height was not measured

# fit a model with all terms
mod.full<-lm(Height~log(DBH)+log(DBH):Species+log(DBH):Year+log(DBH):Species:Year,data=x)

# fit model with only DBH and species terms
mod.species<-lm(Height~log(DBH)+log(DBH):Species,data=x)

# fit model with only DBH and year terms
mod.year<-lm(Height~log(DBH)+log(DBH):Year,data=x)

@@

# Run through these iteratively
anova(mod.full,mod.dbh)

@@

Based on this ANOVA table, there is a significant reduction in residual sum of squares between the full model and the DBH-only model.  The species+DBH and year+DBH models don't seem to add much (not shown).

***
### Compare R^2^ between models
How much variance is explained by each model?  Summary below:

```{r extract r2 and other stats for each model}
rbind(data.frame(Model="full",broom::glance(mod.full)),
      data.frame(Model="dbh-only",broom::glance(mod.dbh)),
      data.frame(Model="dbh+species",broom::glance(mod.species)),
      data.frame(Model="dbh+year",broom::glance(mod.year))

      ) %>% 
  dplyr::select(1:7)

@@

Based on this, and the extremely modest increase in R^2^ from DBH-only to full model, I'm tempted to say that we can pool all species and years into the most parsimonious (ie., DBH:Height) model.  But definitely up for discussion!

### DBH:Height equation for pooled data
If we decide to use pooled data for the DBH:Height equation:

paste("Predicted.Height = ",coef(mod.dbh)[2],
      " x ln(DBH) ",
      coef(mod.dbh)[1])
```

@@

The coefficients using the pooled species/years are pretty similar to Mike's formula for spruce.

>>>>

As per previous discussion, we shouldn't pool measurements across measurement years because this would violate assumption of independence among samples.

****
### Can we pool species?

@@

```{r test for species effect}

# function to compare three models
# separate models will be fit for each year

speciesMod1<-function(year) {

# Assign data
x<-
   dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
  filter(Year==year) %>% 
    drop_na(Height) # remove trees where height was not measured

# fit a model with all terms
mod.full<-lm(Height~log(DBH)+log(DBH):Species,data=x)

@@

# extract coef and other values
out<-data.frame(year=year,intercept=coef(mod.dbh)[1],
                slope=coef(mod.dbh)[2],
                p.value=broom::glance(mod.dbh)$p.value,
                r2=broom::glance(mod.dbh)[1])
rownames(out)=year

# Return
return(list(anov=anova(mod.full,mod.dbh),out=out))

}

@@

Now let's run the model for the four measurement periods:

1992:
```{r}
speciesMod1(1992)$anov
```

***
1994:

```{r}
speciesMod1(1994)$anov
```

***
1997:

```{r}
speciesMod1(1997)$anov
```

***
2019:

```{r}
speciesMod1(2019)$anov
```

For each of the four measurement periods, there is no significant difference between models that include DBH:species interactions and just use DBH alone.  Therefore, we can develop a single height model for each measurement period.

### DBH:Height equation for pooled species data

rbind(speciesMod1(1992)$out,
      speciesMod1(1994)$out,
      speciesMod1(1997)$out,
      speciesMod1(2019)$out) %>% 
  as_tibble()

@@

@@

### 2020-03-21 analysis/prelimAnalysis.Rmd
more updates

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width=8)

@@

pbai<- # assign output so we can plot several periods
  dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
  mutate(BA=(DBH/2/100)^2*pi*20) %>% # convert DBH to basal area and multiply for per hectare
  dplyr::select(BA.Target,Plot,TreeID,Species,Year,BA) %>% 
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA.Sum=sum(BA,na.rm=T)) %>% 
  pivot_wider(names_from=Year,values_from=BA.Sum) %>% 
  mutate(`1992-1994`=`1994`-`1992`) %>% 
  mutate(`1994-1997`=`1997`-`1994`) %>% 
  mutate(`1997-2009`=`2009`-`1997`) %>% 
  mutate(`2009-2019`=`2019`-`2009`)

@@

pbai.Plot(initYear="1994",baPeriod="2009-2019",data=pbai.2,x=5,y=3)

@@

>>>>

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width=10)

@@

pbai<- # assign output so we can plot several periods
  dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
  mutate(BA=(DBH/2/100)^2*pi*20) %>% # convert DBH to basal area and multiply for per hectare
  dplyr::select(BA.Target,Plot,TreeID,Species,Year,BA) %>% 
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA.Sum=sum(BA,na.rm=T)) %>% 
  pivot_wider(names_from=Year,values_from=BA.Sum) %>% 
  mutate(`1992-1994`=`1994`-`1992`) %>% 
  mutate(`1994-1997`=`1997`-`1994`) %>% 
  mutate(`1997-2009`=`2009`-`1997`) %>% 
  mutate(`2009-2019`=`2019`-`2009`) %>% 
  mutate(`1994-2019`=`2019`-`1994`)

@@

pbai.Plot(initYear="2009",baPeriod="2009-2019",data=pbai.2,x=10,y=3)

@@

1994-2019 period:

pbai.Plot(initYear="1994",baPeriod="1994-2019",x=5,y=1)
```

Plot 3 is an outlier because it has a negative BAI over the 1994-2019 period.  With this outlier removed, the slope would likely be p<0.05.
